---
title: "Homework 1"
author: "Denis Ostroushko"
format: pdf
execute: 
  echo: false
  message: false
  warning: false
---

```{r read in packages }
path_main_folder = substr(getwd(), 1, nchar(getwd()) - nchar("HW1"))

source(paste0(path_main_folder, "Master Packages.R"))
```

```{r read data }
timetrial <- read_csv('timetrial.csv')


timetrial_long <- 
  timetrial %>% 
  pivot_longer(
    cols = c("Trial1", "Trial2", "Trial3", "Trial4", "Trial5"), 
    names_to = "Trial", 
    values_to = "Measure"
  ) %>% 
  mutate(
    Trial_num = as.numeric(substr(Trial, nchar("Trial")+1,  nchar("Trial") + 2)
                                  )
  )

```

# Question 1 

## a)

Outcomes for 20th subject are shown in @tbl-y-20

```{r answer 1a }
#| label: tbl-y-20
#| tbl-cap: "Outcomes for 20th subject in the study" 
#| 
timetrial_long %>% 
  filter(Id == '20') %>% 
  select(Measure) %>% 
  kable(booktabs = T, 
        align = 'c', 
      col.names = c("Y20")) %>% 
  kable_styling(full_width = F, latex_options = c("HOLD_position"))
```

## b) 

In the long format of the data, where we have `r length(unique(timetrial_long$Id))` unique participants, and 
`r nrow(timetrial_long)/length(unique(timetrial_long$Id)) %>% round(.)` observations per Id, the full length of responses $\textbf Y$ is `r nrow(timetrial_long)`

## c)

We are given that for a random variable that generate observed values for the 20th participant, $Var(Y_{20 \ j}) =$ 0.05, so 
variance is constant for each observation $j$ = 1, 2, 3, 4, 5. 

Covariance of $i^{th}$ and $j^{th}$ observations is also constant at 0.2. 

Covariance of $i^{th}$ observation with itself is variance, an is 0.05. 

@tbl-vcov-20 displays resulting variance-covariance matrix. 

```{r 1c egenrate vcov matrix}
#| label: tbl-vcov-20
#| tbl-cap: "Variance-covariance matrix for 20th participant" 
vcov_20 <- matrix(c(rep(0.02, 25)), nrow = 5, ncol = 5)

for(i in 1:5){
  vcov_20[i,i] = 0.05
}

vcov_20 %>% 
  kable(booktabs = T
        ) %>% 
  kable_styling(latex_options = "HOLD_position")
```

We can now convert this matrix into the correlation matrix. 

$\rho_{ij} = \frac{Cov(Y_{ij}, Y_{jk}}{\sqrt{Var(Y_{ij}) * Var(Y_{jk})}}$

Since all variances are eaqual tp 0.05, the calculation is pretty straightforward. 

Resulting correlation matrix is shown in @tbl-corr-matrix-20 

```{r correlation matrix 1c }
#| label: tbl-corr-matrix-20 
#| tbl-cap: "Correlation matrix for 20th individual"

(vcov_20 /( sqrt(0.05) *sqrt( 0.05))) %>% 
  kable(booktabs = T
        ) %>% 
  kable_styling(latex_options = "HOLD_position")
  
```

Each observation on the diagonal is equal to 1, which makes the results more credible 

## d) 

A variance-covariance matrix for each individual is $5 \times 5$ in size, and we have 107 individuals, therefore the size of 
full matrix is ($5 \times 107$) $\times$  ($5 \times 107$), which means that the dimension of variance covariance matrix 
${\Sigma}$ is $535 \times 535$. 

## e) 

We consider trial number, age, and sex of a participant as predictors. Values provided in the dataset are given in 
the @tbl-predictors-20. 

```{r}
#| label: tbl-predictors-20
#| tbl-cap: "Covariates for 20th participant" 
#| 
timetrial_long %>% 
  filter(Id == 20)  %>% 
  select(Trial_num, Age, Sex) -> 
  
  x_20 

x_20 %>% 
  kable(
    col.names = c("Trial Number", "Age", "Sex"), 
    booktabs = T
  ) %>% 
  kable_styling(latex_options = "HOLD_position", full_width = F)

```

However, we also need a column of 1s in order to estimate $\hat \beta_0$. A full matrix $\textbf X_{20}$ for the 20th 
participant is given in @tbl-20-full-x

```{r}
#| label: tbl-20-full-x
#| tbl-cap: "Full matrix for model estimation" 
#| 
x_20 %>% 
  cbind(
    data.frame(Intercept = rep(1, 5)), 
    .
  )%>% 
  kable(
    col.names = c(""," ", "", ""), 
    booktabs = T
  ) %>% 
  kable_styling(latex_options = "HOLD_position", full_width = F)
```

## f) 

Including intercept column, the full dimension of $\textbf X$ is 535 $\times$ 4. Without intercept, the size is 
 $\textbf X$ is 535 $\times$ 3. 

# Question 2 

## a) 

```{r 2a boxplot for outliers }
#| label: fig-boxplot-2a
#| fig-cap: "Distbution of Reaction Measuresements in Each Trial"
ggplot(
  data = timetrial_long, 
  aes(x = Trial_num, y = Measure, group = Trial)
) + 
  theme_minimal() + 
  geom_boxplot(fill = NA) + 
  geom_jitter(alpha = 0.5, width = .25) + 
  
  labs(
    title = "Distribtion of Reaction Times", 
    x = "Trial Number", 
    y = "Measurement"
  )

```

```{r 2a spahetti }
#| label: fig-spah-2a
#| fig-cap: "Individual Profiles of Measuresements for each participant"
ggplot(
  data = timetrial_long, 
  aes(x = Trial_num, y = Measure, group = Id)
) + 
  theme_minimal() + 
  geom_line(alpha = 0.5, color = "grey") + 
  stat_smooth(aes(group = 1)) + 
  
  labs(
    title = "Distribtion of Reaction Times", 
    x = "Trial Number", 
    y = "Measurement"
  )
```

## b) 



## c) 

```{r 2c spahetti by age group }

#| label: fig-spah-2a
#| fig-cap: "Individual Profiles of Measuresements for each participant"

timetrial_long <- 
  timetrial_long %>% 
  mutate(
    age_cat = ifelse(Age < 18, "Children", "Adults")
  )

ggplot(
  data = timetrial_long, 
  aes(x = Trial_num, y = Measure, group = Id)
) + 
  theme_minimal() + 
  geom_line(alpha = 0.5, color = "grey") + 
  stat_smooth(aes(group = age_cat, color = age_cat)) + 
  
  labs(
    title = "Distribtion of Reaction Times", 
    x = "Trial Number", 
    y = "Measurement", 
    color = ""
  )

```

# Question 3 

## a) 

```{r correlation matrix viz 3a }
#| label: fig-corrplot-3a
ggcorrplot(
  timetrial %>% select(Trial1, Trial2, Trial3, Trial4, Trial5) %>% cor(),
  type = "upper",
  outline.color = "white",
  lab = TRUE
)

```

## b) 

# Question 4 

## a) 

```{r 4a t test}

timetrial_long %>% 
  group_by(Trial) %>% 
  summarize(mean(Measure))

t.test(
  x = timetrial %>% select(Trial5) %>% unlist() , 
  y = timetrial %>% select(Trial1) %>% unlist() ,
  alternative = "less"
)

```

## b) 

### i) 

### ii) 

### iii) 

