---
title: "Analysis"
format: pdf
execute: 
  echo: false
  message: false
  warning: false
---

```{r}
#| results: hide
#| include: false 
path_main_folder = substr(getwd(), 1, nchar(getwd()) - nchar("Final Project"))

source(paste0(path_main_folder, "Master Packages.R"))

```

```{r read and modify data}

curran <- read_csv("CurranLong.csv", show_col_types = F)
curran <- curran %>% mutate(momage_tv = momage + occasion * 2)
# id: child id - our cluster id 

# Predictors in a model 
  # kid gen - presumed kid sex: 0 = F, 1 = M - cluster invariant 
  # mom age - cluster invariant 
  # homecog - measure of home cognitive stimulation: seems cluster invariant (measured at time = 1, presumed baseline )
  # homeemo - same 
  # occasion - identify time points - cluster variant 
  # kidagetv - time variant age measure: recorded for each new follow up measure time point 

# outcomes: 
  # read - primary outcome 
  # anti - measure of anti social behavior - secondary outcome 
```

### Statistical Analysis 

# Results 

### Study Population 

```{r descriptive stats table one }
#| results: hide 

CreateTableOne(
  data = curran, 
  strata = "occasion", 
  vars = c('read', 'anti', 'kidage', 'kidgen', 'momage_tv', 'homecog', 'homeemo'), 
  factorVars = c('kidgen'),
  
  includeNA = T, 
  test = F, 
  smd = F, 
  addOverall = T
  
) -> table1

print(table1, smd = TRUE, showAllLevels = F, varLabels = TRUE) -> table1_for_kable


```

```{r}
#| label: tbl-table-one
#| tbl-cap: "Population summary at each follow up time point" 

rownames(table1_for_kable) <- c("N", "Reading Score (SD)", "Antisocial Score (SD)", "Child Age (SD)", "% Male",
                                "Mom's Age", "Congitive Stimulation (SD)", "Emotional Stimulation (SD)")

nr <- nrow(table1_for_kable)

table1_for_kable %>% 
  kable(
    linesep = "\\addlinespace", 
    col.names = c("Overall", "Baseline", "First Follow Up",  "Second Follow Up",  "Third Follow Up"), 
    booktabs = T,
    longtable = TRUE, 
    align = 'c'
    ) %>% 
  kable_styling(
    latex_options = c("striped", "hover", "condensed", "HOLD_position"), 
    font_size = 9
    )  %>% 
  column_spec(2, bold = T) %>% 
  # pack_rows(group_label = "Outcome Measures", start_row = 1, end_row = 2) %>% 
  # pack_rows(group_label = "Predictors", start_row = 3, end_row = nr) %>% 
  add_footnote("Cognitive and Emotional Stimulation Scores were recorded at baseline and do not vary with time") 
  

```

```{r }
#| label: fig-prim-outliers-plot-2
#| eval: false 
#| fig-cap: "Some outliers in the early observation periods" 
#| fig-width: 12
#| fig-height: 8
grid.arrange(
  ggplot(curran, 
         aes(x = read, 
             group = as.factor(occasion), 
             fill = as.factor(occasion)
             )) + geom_histogram() + theme_minimal() + 
    theme(legend.position = "bottom"), 
  ggplot(curran, 
         aes(y = read, 
             group = as.factor(occasion), 
             fill = as.factor(occasion)
             )) + geom_boxplot(outlier.color = "red") + theme_minimal() + 
    scale_x_discrete(labels = c(0,1,2,3)) + 
    theme(legend.position = "bottom"), 
  nrow = 1
)

```

```{r not-evaluated primary and secondary model development steps }
#| eval: false 

fullest_model_time_v1 <- 
  geeglm(read ~ (homecog + homeemo + kidgen + kidage + occasion)^2, 
       data = na.omit(curran), 
       id = id, 
       corstr = "exchangeable", 
       family = gaussian(link = "identity"))

anova(fullest_model_time_v1) %>% tidy() %>% filter(p.value < 0.05) %>% 
  select(term) %>% unlist() %>% 
  paste(., collapse = " + ") -> linear_predictor

model_formula <- as.formula(
  paste("read ~ kidgen +", linear_predictor,  " + homeemo:homecog")
)

primary_analysis_model <- 
  geeglm(model_formula, 
       data = na.omit(curran), 
       id = id, 
       corstr = "exchangeable", 
       family = gaussian(link = "identity"))

#################################
# compare standard errors and efficiency of data use with and without homeemo:homecog interaction 

primary_analysis_model %>% tidy() %>% select(term, estimate, std.error) %>% 
  rename(estimate_no_int = estimate, 
         str.error_no_int = std.error) -> no_int

model_formula2 <- as.formula(
  paste("read ~ kidgen +", linear_predictor, " + homeemo:homecog")
)

primary_analysis_model2 <- 
  geeglm(model_formula2, 
       data = curran, 
       id = id, 
       corstr = "exchangeable", 
       family = gaussian(link = "identity"))

primary_analysis_model2 %>% tidy() %>% select(term, estimate, std.error) %>% 
  rename(estimate_int = estimate, 
         str.error_int = std.error) -> int

comp_ <- int %>% 
  left_join(no_int, by = "term")

comp_ %>% 
  mutate(diff = str.error_int - str.error_no_int)

############
# store results 

write_rds(primary_analysis_model, "primary model.rds")


fullest_model_time_v1 <- 
  geeglm(anti ~ (homecog + homeemo + kidgen + kidage + occasion)^2, 
       data = curran, 
       id = id, 
       corstr = "exchangeable", 
       family = gaussian(link = "identity"))

anova(fullest_model_time_v1) 

anova(fullest_model_time_v1) %>% tidy() %>% filter(p.value < 0.05) %>% 
  select(term) %>% unlist() %>% 
  paste(., collapse = " + ") -> linear_predictor

model_formula <- as.formula(
  paste("anti ~ kidgen + kidage + ", linear_predictor,  " + homeemo:homecog")
)


secondary_analysis_model <- 
  geeglm(model_formula, 
       data = curran, 
       id = id, 
       corstr = "exchangeable", 
       family = gaussian(link = "identity")) 

secondary_analysis_model %>% tidy() %>% mutate(p.value = round(p.value, 4), 
                                               sognif = ifelse(p.value < 0.05, "*", ""))

write_rds(secondary_analysis_model, "secondary model.rds")
  
```


## Cognitive and Emotional Stimulation Measures 

### Reading Comprehension Scores 

<!-- 
Original text 

@tbl-prim-model-est shows  main and interaction effects of at home emotional and cognitive stimulation scores on the reading comprehension scores, as well as other predictors from the Gaussian GEE with exchangeable correlation structure. We found that the effect of emotional and cognitive scores varies at each time point. Effect modification associated with each additional follow up time point for cognitive scores was 0.01 (0 to 0.03 95% normal confidence interval), suggesting that as children got older, the effect of cognitive stimulation on reading scores increased more (P=0.16). However, the results are not statistically related to the outcome at the pre-determined significance level $\alpha = 0.05$. Similarly, the effect of 
emotional at home stimulation increased even greater. Effect modification was 0.03 (0.01 to 0.05 95% confidence interval), indicating that emotional stimulation score recorded at baseline amplified with extra additional observational time (P<0.001).

The main effects of emotional and cognitive stimulation measure the strength of association between these scores and reading 
ability at baseline. We found that with one unit increase in the cognitive score, reading ability increased by 0.08 (-0.01 to 0.18 95% normal confidence interval), indicating some effect of cognitive scores in reading abilities, but the results are 
more suggestive rather than statistically definitive (P = 0.08). The effect of emotional stimulation scores on child's reading ability at baseline was 0.05. Each additional unit on the emotional stimulation scale resulted in the 0.05 (-0.03 to 0.13 95% confidence interval) increase in reading comprehension scores. This effect was not strongly associated with the outcome variable (P=0.24), implying that other considered predictors are stronger at explaining variation in the reading scores. 

Model estimate for interaction of the emotional and cognitive scores at baseline was estimated at 0.00 (-0.01 to 0.01 95% normal confidence interval), suggesting no evidence that the effect of one predictor varies for different levels of the other predictor (P=0.38). 


REFORMAT TEXT: 
--> 

@tbl-prim-model-est displays the main and interaction effects of at-home emotional and cognitive stimulation scores on reading comprehension scores, as well as estimates for other predictors from the Gaussian GEE model with an exchangeable correlation structure. We observed variation in the impact of emotional and cognitive scores at each follow-up time point. The effect modification per additional follow-up for cognitive scores was 0.01 (95% CI: 0 to 0.03) after adjusting for other predictors, indicating that as children aged, the influence of cognitive stimulation on reading scores intensified, albeit not significantly (P=0.16). Comparatively, the role of emotional stimulation escalated more prominently over time, with an effect modification of 0.03 (95% CI: 0.01 to 0.05)  after adjusting for other predictors, denoting that the emotional stimulation score amplified with each extra follow-up (P<0.001).

The main effects of emotional and cognitive stimulation quantify the baseline association between these scores and reading proficiency. A one-unit increase in the cognitive score corresponded to a 0.08 (-0.01 to 0.18 95% CI) uptick in reading ability  after adjusting for other predictors, hinting at some effect of cognitive scores on literacy skills, though not conclusively (P=0.08). The emotional stimulation score's baseline impact on reading comprehension was 0.05; every added emotional stimulation unit yielded a 0.05 (-0.03 to 0.13 95% CI) rise in scores  after adjusting for other predictors, an influence not strongly linked to variation in reading outcomes (P=0.24). This suggests other modeled predictors better elucidate reading score variability.

The estimated interaction between emotional and cognitive scores at baseline was 0.00 (-0.01 to 0.01 95% CI)  after adjusting for other predictors, offering no evidence that one predictor's effect changes across levels of the other (P=0.38).

```{r primary outcome model }
#| label: tbl-prim-model-est
#| tbl-cap: "Gaussian GEE Main Effects on Reading Comprehension Score" 
primary_analysis_model <- readRDS("primary model.rds")

primary_analysis_model %>% tidy(conf.int = TRUE)  %>% 
  mutate(
    term_new  = c("(Intercept)", 		
                  "Sex Male",				
                  "Cog. Score", 				
                  'Emo. Score',				
                  'Age',				
                  'Follow Up Period',				
                  'Cog. Score * Follow Up Period',				
                  'Emo. Score * Follow Up Period',				
                  'Age * Follow Up Period',				
                  'Cog. Score * Emo. Score'
                  ), 
    ci = paste0("(", 
                round(conf.low, 2), 
                ", ", 
                round(conf.high, 2), 
                ")")
  )  %>% 
  select(term_new, estimate, ci, p.value) %>% 
  kable(digits = 2, 
        linesep = "\\addlinespace", 
        align = c('l', 'c', 'c', 'c'), 
        col.names = c("Model Term", "Estimate", "95% CI", "P-value"), 
        booktabs = T) %>% 
  kable_styling(
    latex_options = c("striped", "hover", "condensed", "HOLD_position")
    ) %>% 
  add_footnote("Terms with '*' denote interactions") %>% 
  add_footnote("Cog. Score = Cognitive Stimulation at Home Score") %>% 
  add_footnote("Emo. Score = Emotional Stimulation at Home Score")

```
<!-- 
Original text: 

Variation in reading scores is greatly explained by follow up period and age of a child at baseline.
On average, reading scores improved by 2.03 (1.55 to 2.5 95% normal confidence interval) points with each follow up 
period, verifying that as children age, they statistically get better at reading (P < 0.001). Additionally, on average, older children at baseline had higher reading scores (P < 0.001), as on extra year of age at baseline resulted in about 0.8 (0.67 to 0.93 95% confidence interval) increase in reading score. 

However, the effect of age diminished at each additional follow up period. We estimate that the effect of age decreased by 
0.19 [-0.19 bounded by (-0.25 to -0.12) 95% confidence interval] at each additional follow up period. There is strong statistical evidence (P < 0.001)  
that as children grow older from ages between six and eight into early teenage years, initial observed differences in reading scores vanish over time. Visually, observed on @fig-prim-outliers-plot-2 that the spread of reading scores decreased at 
additional follow up point.  

REWORK: 
--> 
Reading score variability was largely explained by the follow-up period and child age at baseline. On average, reading scores rose by 2.03 (95% CI: 1.55 to 2.50) points per follow-up interval  after adjusting for other predictors, affirming that literacy skills statistically improved with age over time (P<0.001). Additionally, older children initially showed markedly higher reading achievement (P<0.001); an extra baseline year corresponded with a 0.80 (95% CI: 0.67 to 0.93) increase in scores  after adjusting for other predictors.

<!-- 
However, the influence of age waned at each successive follow-up. We estimate a 0.19 (95% CI: -0.25 to -0.12) decrease in the age effect at every additional follow-up, denoting strong evidence (P<0.001) that initial reading score differences across ages 6 to 8 years vanished in later assessments during early adolescence. This pattern is visually evident in @fig-prim-outliers-plot-2, which exhibits narrowed reading score dispersion at subsequent time points.

We further investigate main and interaction effects of age, time, emotional, and cognitive scores using regression effect plots. 
@fig-effect-time-by-age compares estimated reading scores over time for children with varying baseline age. We select values 
based on study inclusion criteria, and present learning effect curves obtained using Gaussuan GEE with exchangeable correlation structure. It is evident that the greatest difference in the average reading scores is observed at baseline, and the youngest 
children have the largest average learning rate, or the largest increase in reading scores. By the time of the third follow up point, which took place 6 years after the baseline measurements, children of all starting ages seem to have more similar 
average reading scores. 

--> 

However, the influence of age diminished at each successive follow-up. We estimate a 0.19 (95% CI: -0.25 to -0.12) decrease in the age effect at every additional follow-up  after adjusting for other predictors, denoting strong evidence (P<0.001) that initial reading score differences across ages 6 to 8 years vanished in later assessments during early adolescence. 

We further investigated the main and interaction effects of age, time, emotional, and cognitive scores using regression effect plots. Figure @fig-effect-time-by-age presents estimated reading score trajectories over time for children of varying baseline ages, selected per study inclusion criteria. The learning effect curves were obtained via Gaussian GEE with an exchangeable correlation structure. The graph highlights that the greatest difference in mean reading scores occurs at baseline, with the youngest children displaying the steepest growth in reading skills on average. By the third follow-up roughly 6 years after baseline measurements, children across all starting ages converge to more similar average reading score levels.

```{r}
#| label: fig-effect-time-by-age
#| fig-cap: "Blank"
#| fig-height: 3
#| fig-width: 5
#| 
# main effect main model for interactions 

# quantile(curran$kidage, c(0.25, 0.5, 0.75))

plot_model(model = primary_analysis_model, 
           type = "eff", 
           terms = c("occasion", "kidage [6, 7, 8]")) + 
  theme_minimal() + 
  labs(x = "Follow Up Period", 
       y = "Reading Score", 
       color = "Child's Age at Baseline") + 
  theme(legend.position = "bottom", 
        legend.text = element_text(size = 8)) + 
  ggtitle("Effect of Time on Reading Scores")
```

```{r not evaluated - numbers for text}
#| eval: false 
curran %>% 
  group_by(homeemo) %>% 
  summarise(ids = length(unique(id)), 
            p_total = length(unique(id))/length(unique(curran$id)))

curran %>% 
  group_by(homecog) %>% 
  summarise(ids = length(unique(id)), 
            p_total = length(unique(id))/length(unique(curran$id)))

summary(primary_analysis_model)
vcov(primary_analysis_model)

cog_adj_slope <- function(cog) {2.026182  + 0.01166 * cog }
var_cog_adj_slope <- function(cog) {5.850e-02  + 7.020e-05 * cog^2 + 2 * cog * (-1.598e-04) }

cog_adj_slope(7)

cog_adj_slope(7) - 1.96 * sqrt(var_cog_adj_slope(7))
cog_adj_slope(7) + 1.96 * sqrt(var_cog_adj_slope(7))

cog_adj_slope(9)

cog_adj_slope(9) - 1.96 * sqrt(var_cog_adj_slope(9))
cog_adj_slope(9) + 1.96 * sqrt(var_cog_adj_slope(9))


cog_adj_slope(11)

cog_adj_slope(11) - 1.96 * sqrt(var_cog_adj_slope(11))
cog_adj_slope(11) + 1.96 * sqrt(var_cog_adj_slope(11))

summary(primary_analysis_model)
vcov(primary_analysis_model)

emo_adj_slope <- function(emo) {2.026182  + 0.02853 * emo }
var_emo_adj_slope <- function(emo) {5.850e-02  + 7.17e-05 * emo^2 + 2 * emo * (-5.83e-04) }

emo_adj_slope(8)
emo_adj_slope(8) - 1.96 * sqrt(var_emo_adj_slope(8))
emo_adj_slope(8) + 1.96 * sqrt(var_emo_adj_slope(8))

emo_adj_slope(10)
emo_adj_slope(10) - 1.96 * sqrt(var_emo_adj_slope(10))
emo_adj_slope(10) + 1.96 * sqrt(var_emo_adj_slope(10))

emo_adj_slope(12)
emo_adj_slope(12) - 1.96 * sqrt(var_emo_adj_slope(12))
emo_adj_slope(12) + 1.96 * sqrt(var_emo_adj_slope(12))

```
<!-- 
FIGURE 1 visualizes interactions between stimulation scores and learning rates over time. We selected sample 25th, 50th, and 75th
quantiles of cognitive and emotional stimulation scores to show the effect of baseline score on the learning rates. 

For an average child who had a cognitive stimulation score of 7 at baseline, corresponding to 25th percentile of all observed scores, expected two year change in the reading scores was 2.11 (1.63 to 2.59 95% CI). For comparison, the rate of change for an average child with a score of 9 at baseline was 2.13 (1.65 to 2.62 95% CI). An average child with a cognitive stimulation score of 11 had an average change in reading score of 2.15 (1.66 to 2.65 95% CI). These results might be suggestive that cognitive stimulation at baseline is improving the rate at which children learn to read, however, further research and data collection are required to make more definitive statements. 


--> 

@fig-main-effects-by-age shows interactions between stimulation scores and reading skill growth rates over time. We chose sample $25^{th}$, $50^{th}$, and $75^{th}$ quantiles of cognitive and emotional stimulation scores to display the effect of baseline scores on learning rates.

For an average child with a baseline cognitive score at the $25^{th}$ percentile (score of 7), the expected two-year change in reading scores was 2.11 (95% CI 1.63 to 2.59). In comparison, the rate of change for an average child with a baseline score of 9 was 2.13 (95% CI 1.65 to 2.62). An average child with a cognitive score of 11 had an average reading score change of 2.15 (95% CI 1.66 to 2.65). All estimates are marginal effects after adjusting for other variables. These results suggest that higher cognitive stimulation at baseline may improve children reading skill development rates, though more research and data are needed to make definitive conclusions.

Similarly, we adjust the effect of time on reading scores using $25^{th}$, $50^{th}$, and $75^{th}$ percentiles of emotional scores in the study population. the marginal effect of one additional follow up period on reading literacy for a child with a baseline emotional stimulation score of 8 is 2.25, bounded by the (1.8 to 2.71) 95% confidence interval. 
The rate of reading scores change for children with emotional scores of 10 and 12 are 2.31 and 2.37, bounded by (1.86 to 2.77) and (1.91) to (2.83) respectively. All estimates are marginal and summarize the data while adjusting for other predictors in the model. 

```{r}
#| label: fig-main-effects-by-age
#| fig-cap: "Values of Emotional and Cognitive Scores are chosen at 25th, 50th, and 75th sample percentiles" 
#| fig-height: 4
#| fig-width: 9
# main effect main model for interactions 

# quantile(curran$homecog, c(0.25, 0.5, 0.75))
# quantile(curran$homeemo, c(0.25, 0.5, 0.75))

# summary(curran$homecog)
# summary(curran$homeemo)
plot_model(model = primary_analysis_model, 
           type = "eff", 
           terms = c("occasion", "homecog [7, 9, 11]")) + 
  theme_minimal() + 
  labs(x = "Follow Up Period", 
       y = "Reading Score", 
       color = "Cognitive Stimulation Score") + 
  ggtitle("") + 
  theme(legend.position = "bottom", 
        legend.text = element_text(size = 8)) -> p_left


plot_model(model = primary_analysis_model, 
           type = "eff", 
           terms = c("occasion", "homeemo [8, 10, 12]")) + 
  theme_minimal() + 
  labs(x = "Follow Up Period", 
       y = "Reading Score", 
       color = "Emotional Stimulation Score") + 
  ggtitle("") + 
  theme(legend.position = "bottom", 
        legend.text = element_text(size = 8)) -> p_right

grid.arrange(p_left, p_right, nrow = 1)

```

### Antisocial Scores 

```{r secondary outcome model }
#| label: tbl-sec-model-est
#| tbl-cap: "Gaussian GEE Main Effects on Antisocial Behavioral Score" 
#| 
secondary_analysis_model <- readRDS("secondary model.rds")

secondary_analysis_model %>% tidy(conf.int = TRUE)  %>% 
  mutate(
    term_new  = c('(Intercept)'				,
                  'Sex Male'				,
                  "Age", 
                  'Cog. Score'				,
                  'Emo. Score'				,
                  'Follow Up Period'				,
                  'Follow Up Period * Age'	,		
                  'Cog. Score * Emo. Score'
                  ), 
    ci = paste0("(", 
                round(conf.low, 2), 
                ", ", 
                round(conf.high, 2), 
                ")")
  )  %>% 
  select(term_new, estimate, ci, p.value) %>% 
  kable(digits = 2, 
        linesep = "\\addlinespace", 
        align = c('l', 'c', 'c', 'c'), 
        col.names = c("Model Term", "Estimate", "95% CI", "P-value"), 
        booktabs = T) %>%
  kable_styling(
    latex_options = c("striped", "hover", "condensed", "HOLD_position")
    ) %>% 
  add_footnote("Terms with '*' denote interactions") %>% 
  add_footnote("Cog. Score = Cognitive Stimulation at Home Score") %>% 
  add_footnote("Emo. Score = Emotional Stimulation at Home Score")

```

```{r}
#| label: fig-effect-time-by-age-sec
#| fig-cap: "25th Percentile: 6.3 Years old. 50h Percentile: 6.8 Years old. 75th Percentile: 7.5 Years old. "
#| 
# main effect main model for interactions 

# quantile(curran$kidage, c(0.25, 0.5, 0.75))

plot_model(model = secondary_analysis_model, 
           type = "eff", 
           terms = c("occasion", "kidage [6.33, 6.83, 7.50]")) + 
  theme_minimal() + 
  labs(x = "Follow Up Period", 
       y = "Antisocial Score", 
       color = "Child's Age at Baseline") + 
  theme(legend.position = "bottom", 
        legend.text = element_text(size = 8)) + 
  ggtitle("Effect of Time on Antisocial Scores")
```

```{r}
#| label: fig-sec-effects-by-age
#| fig-cap: "Values of Emotional and Cognitive Scores are chosen at 25th, 50th, and 75th sample percentiles" 
#| fig-height: 6
#| fig-width: 9
# main effect main model for interactions 

# quantile(curran$homecog, c(0.25, 0.5, 0.75))

plot_model(model = secondary_analysis_model, 
           type = "eff", 
           terms = c("occasion", "homecog [7, 9, 11]")) + 
  theme_minimal() + 
  labs(x = "Follow Up Period", 
       y = "Antisocial Score", 
       color = "Cognitive Stimulation Score") + 
  ggtitle("") + 
  theme(legend.position = "bottom", 
        legend.text = element_text(size = 8)) -> p_left


plot_model(model = secondary_analysis_model, 
           type = "eff", 
           terms = c("occasion", "homeemo [8, 10, 12]")) + 
  theme_minimal() + 
  labs(x = "Follow Up Period", 
       y = "Antisocial Score", 
       color = "Emotional Stimulation Score") + 
  ggtitle("") + 
  theme(legend.position = "bottom", 
        legend.text = element_text(size = 8)) -> p_right

grid.arrange(p_left, p_right, nrow = 1)

```

# Discussion 