---
title: "Homework 0"
format: pdf
execute: 
  echo: false
  message: false
  warning: false
---

```{r}
library(tidyverse)
library(kableExtra)
library(data.table)
```

```{r}
#| label: read-data 

long_chicks <- read_csv('chicks.csv')

long_chicks <- data.table(long_chicks)

long_chicks <- long_chicks[, new_variable := (seq_len(.N)-1) * 2, by = Chick]

long_chicks <- 
  long_chicks %>% 
  mutate(
    new_variable = 
      case_when(
        Time == 21 ~ new_variable - 1,
        T ~ new_variable
      ), 
    new_variable
  )

long_chicks %>% 
  filter(new_variable != Time) %>% nrow() -> missing_obs_count

```

# Brief Data Set Sumamry: 

* Number of unique chicks: `r long_chicks %>% select(Chick) %>% unique %>% nrow`

```{r}
#| label: count-missing-data

long_chicks %>% 
  mutate(
    max = max(Time)
  ) %>% 
  filter(
    Time == max
  ) %>% 
  select(Chick) %>% 
  unique() %>% 
  unlist() %>% 
  length() -> complete_obs

long_chicks %>% 
  select(Chick) %>% 
  unique() %>% nrow() -> total

long_chicks %>% 
  group_by(Chick) %>% 
  summarise(
    n = n(), 
    last_obs = max(Time)
  ) %>% 
  
  filter(
    
    n != 
      length(
        seq(from = 0, to = 20, by = 2) 
      ) + 1
  ) -> last_time_for_inc
```

* Total observations: `r long_chicks %>% nrow`

* Average number of observations per Chick: `r (long_chicks %>% nrow)/(long_chicks %>% select(Chick) %>% unique %>% nrow) %>% round(.,2)`; 12 observations for those wtih complete data

* Number of Chicks with complete observation period: `r complete_obs`; number with last time before time = 21: `r total - complete_obs`

* Number of Chicks with skipped measurement: `r missing_obs_count`

* Number of observations for Chicks with incomlete data (maximum number of observation periods is 12): 

```{r}
last_time_for_inc %>% 
  kable(booktabs = T, 
        align = 'c',
        col.names = c("Chick ID", "N observed Periods", "Last Observed Time")
        ) %>% 
  kable_styling(
    latex_options = "striped"
  )
```

# Longitudinal Plot

```{r}

max <- 
  long_chicks %>% 
  summarise(m  = max(weight)) %>% 
  unlist()

max_for_plot <- 
  ceiling(
    max / 10^(nchar(max)-1) 
  ) * 10^(nchar(max)-1) 

long_chicks %>% 
  ggplot(
    aes(
      x = Time, 
      y = weight, 
      group = Chick, 
      color = as.factor(Diet)
    )
  ) + 
  geom_line(alpha = 0.25) + 
  geom_smooth(
    aes(group = as.factor(Diet))
  ) + 
  theme_minimal() + 
  theme(legend.position="bottom",
        plot.title = element_text(hjust = 0.5)) +  # calling theme here after theme_minimal allows to fuether customization 
                                      # calling theme minimal after this block overwrites everything we define in theme() 
  labs(
    title = "Average Weigth Gain By Diet Type", 
    x = "Time", 
    y = "Weigth", 
    color = "Diet Type"
  ) + 
  scale_x_continuous(
    breaks = seq(from = 0, to = 21, by  = 3) 
  ) + 
  scale_y_continuous(
    limits = c(0, max_for_plot), 
    breaks = seq(from = 0, 
                 to = max_for_plot,
                 by = 100)
  ) 

```

# Regression Analysis 

```{r}

lm <- lm(weight ~ as.factor(Diet) * Time, data = long_chicks)

summary(lm)

```