---
title: "Final Project Data Exploration"
author: "Denis Ostroushko"
format: pdf
execute: 
  echo: false
  message: false
  warning: false
---

```{r read in packages }
#| results: hide
#| include: false 
path_main_folder = substr(getwd(), 1, nchar(getwd()) - nchar("Final Project"))

source(paste0(path_main_folder, "Master Packages.R"))
```

```{r read and modify data}

curran <- read_csv("CurranLong.csv", show_col_types = F)

# id: child id - our cluster id 

# Predictors in a model 
  # kid gen - presumed kid sex: 0 = F, 1 = M - cluster invariant 
  # mom age - cluster invariant 
  # homecog - measure of home cognitive stimulation: seems cluster invariant (measured at time = 1, presumed baseline )
  # homeemo - same 
  # occasion - identify time points - cluster variant 
  # kidagetv - time variant age measure: recorded for each new follow up measure time point 

# outcomes: 
  # read - primary outcome 
  # anti - measure of anti social behavior - secondary outocme 
```

```{r verify cluster invariate-ness}

curran %>% 
  group_by(id) %>% 
  summarise(N = n(), 
            sex_c = length(unique(kidgen)), 
            homecog_c = length(unique(homecog)), 
            homeemo_c = length(unique(homeemo)),
            
            occ_c = length(unique(occasion)),
            age_c = length(unique(kidagetv))
            ) -> data_sum

print("Distribution of Available time Points per subject")
summary(data_sum$N)
print("Most Subjects >= 3 time points, but data shows some imbalance")

print("Distribution of Unique Counts of Homecog - confirmed cluster invariant variable")
table(data_sum$homecog_c)

print("Distribution of Unique Counts of Homeemo - confirmed cluster invariant variable")
table(data_sum$homeemo_c)

```

# Primary Outcome 

@fig-profile  shows: 

  * Linear Trend in Reading ability against time 
  * High degree of variation overall, variation increases as time goes on 

```{r examine overall variabtion in the population for trend against follow up tme and kids ages}
#| label: fig-profile 
#| fig-cap: "Trend of Reading ability vs Follow Up Time and Kid's age" 
#| fig-height: 8
#| fig-width: 12
grid.arrange(

  ggplot(data = curran, 
         mapping = aes(x = occasion, y = read)) + 
    geom_point(aes(color = kidagetv)) + 
    stat_smooth(method = "lm") + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Reading Score", 
         color = "Kid's age"), 

  
  ggplot(data = curran, 
         mapping = aes(x = kidagetv, y = read)) + 
    geom_point(aes(color = as.factor(occasion))) + 
    stat_smooth(method = "lm", color = "black", linetype = "dashed") +
    stat_smooth(color = "black") +  
    labs(x = "Kid's age", 
         y = "Reading Score", 
         color = "Follow up Time") + 
    theme(legend.position = "bottom"), 
  
  nrow = 1
)

```

@fig-profile-2 shows: 

  * Likely a strong linear trend persists 
  * Individual effect counts for a large portion of variation 
  * Individual variation is strong in a setting with both time variant predictors 

```{r examine individual variabtion in the population for trend against follow up tme and kids ages}
#| label: fig-profile-2 
#| fig-cap: "Trend of Reading ability vs Follow Up Time and Kid's age for Individuals" 
#| fig-height: 8
#| fig-width: 12
grid.arrange(

  ggplot(data = curran, 
         mapping = aes(x = occasion, y = read, group = id)) + 
    geom_line(alpha = 0.5, aes(color = id)) + 
    stat_smooth(method = "lm", linetype = "dashed", color = "black", aes(group = 1)) + 
    stat_smooth(linetype = "dashed", color = "black", aes(group = 1)) + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Reading Score", 
         color = "ID of subject"), 

  ggplot(data = curran, 
         mapping = aes(x = kidagetv, y = read, group = id)) + 
    geom_line(alpha = 0.5, aes(color = id)) + 
    stat_smooth(method = "lm", color = "black", aes(group = 1)) + 
    stat_smooth(linetype = "dashed", color = "black", aes(group = 1)) + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Reading Score", 
         color = "ID of subject"), 
  
  nrow = 1
)

```

@fig-profile-3 shows trend in reading ability by mom's age. Seems that older mom's, on average, have greater intercept for reading ability, but similar slope 

```{r examine individual variabtion in the population for trend against follow up tme and kids ages by moms age}
#| label: fig-profile-3
#| fig-cap: "Trend of Reading ability vs Follow Up Time and Kid's age for Individuals" 
#| fig-height: 8
#| fig-width: 12
grid.arrange(

  ggplot(data = curran, 
         mapping = aes(x = occasion, y = read, group = id)) + 
    geom_line(alpha = 0.5, aes(color = momage)) + 
    stat_smooth(method = "lm", linetype = "dashed", color = "black", aes(group = 1)) + 
    stat_smooth(linetype = "dashed", color = "black", aes(group = 1)) + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Reading Score", 
         color = "ID of subject"), 

  ggplot(data = curran, 
         mapping = aes(x = kidagetv, y = read, group = id)) + 
    geom_line(alpha = 0.5, aes(color = momage)) + 
    stat_smooth(method = "lm", color = "black", aes(group = 1)) + 
    stat_smooth(linetype = "dashed", color = "black", aes(group = 1)) + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Reading Score", 
         color = "ID of subject"), 
  
  nrow = 1
)

```

@fig-profile-homeecog: hard to determine effect of `homecog` on the reading ability 

```{r examine individual variabtion in the population for trend against follow up tme and kids ages}
#| label: fig-profile-homeecog
#| fig-cap: "Trend of Reading ability by Homecog Score" 
#| fig-height: 8
#| fig-width: 12
grid.arrange(

  ggplot(data = curran, 
         mapping = aes(x = occasion, y = read, group = id)) + 
    geom_line(alpha = 0.5, aes(color = homecog)) + 
    stat_smooth(method = "lm", linetype = "dashed", color = "black", aes(group = 1)) + 
    stat_smooth(linetype = "dashed", color = "black", aes(group = 1)) + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Reading Score", 
         color = "Home Cog Score"), 

  ggplot(data = curran, 
         mapping = aes(x = read, group = homecog, fill = homecog)) + 
    geom_histogram() + 
    scale_fill_gradientn(colors = rainbow(5)) , 
  
  nrow = 1
)

```


```{r}
#| label: fig-home-predictors
#| fig-cap: "Relationship between Baseline Home variables and Reading ability" 
#| fig-height: 8
#| fig-width: 12
grid.arrange(
  ggplot(data = curran, 
       aes(x = homecog, y = read)) + 
  geom_point(aes(color = as.factor(occasion))) + 
  geom_smooth() + 
  theme(legend.position = "bottom"), 
  
  ggplot(data = curran, 
       aes(x = homeemo, y = read)) + 
  geom_point(aes(color = as.factor(occasion))) + 
  geom_smooth() + 
  theme(legend.position = "bottom"),
  
  nrow = 1
)
```

# Secondary Outcome 

transforming anti-social behavior to the logarithmic scale 

```{r}
curran$log_anti <- log(curran$anti + 1)
```

```{r }

#| fig-height: 8
#| fig-width: 12
grid.arrange(

  ggplot(data = curran, 
         mapping = aes(x = anti, group = homecog, fill = homecog)) + 
    geom_histogram() + 
    scale_fill_gradientn(colors = rainbow(5)) , 

  ggplot(data = curran, 
         mapping = aes(x = log_anti, group = homecog, fill = homecog)) + 
    geom_histogram() + 
    scale_fill_gradientn(colors = rainbow(5)) , 
  
  nrow = 1
)

```


```{r }
 
#| fig-height: 8
#| fig-width: 12
grid.arrange(

  ggplot(data = curran, 
         mapping = aes(x = occasion, y = log_anti)) + 
    geom_point(aes(color = kidagetv)) + 
    stat_smooth(method = "lm") + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Antisocial Score", 
         color = "Kid's age"), 

  ggplot(data = curran, 
         mapping = aes(x = kidagetv, y = log_anti)) + 
    geom_point(aes(color = as.factor(occasion))) + 
    stat_smooth(method = "lm", color = "black", linetype = "dashed") +
    stat_smooth(color = "black") +  
    labs(x = "Kid's age", 
         y = "Antisocial Score", 
         color = "Follow up Time") + 
    theme(legend.position = "bottom"), 
  
  nrow = 1
)

```


```{r }

#| fig-height: 8
#| fig-width: 12
grid.arrange(

  ggplot(data = curran, 
         mapping = aes(x = occasion, y = log_anti, group = id)) + 
    geom_line(alpha = 0.5, aes(color = id)) + 
    stat_smooth(method = "lm", linetype = "dashed", color = "black", aes(group = 1)) + 
    stat_smooth(linetype = "dashed", color = "black", aes(group = 1)) + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Antisocial Score", 
         color = "ID of subject"), 

  ggplot(data = curran, 
         mapping = aes(x = kidagetv, y = log_anti, group = id)) + 
    geom_line(alpha = 0.5, aes(color = id)) + 
    stat_smooth(method = "lm", color = "black", aes(group = 1)) + 
    stat_smooth(linetype = "dashed", color = "black", aes(group = 1)) + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Antisocial Score", 
         color = "ID of subject"), 
  
  nrow = 1
)

```

```{r }

#| fig-height: 8
#| fig-width: 12
grid.arrange(

  ggplot(data = curran, 
         mapping = aes(x = occasion, y = log_anti, group = id)) + 
    geom_line(alpha = 0.5, aes(color = homecog)) + 
    stat_smooth(method = "lm", linetype = "dashed", color = "black", aes(group = 1)) + 
    stat_smooth(linetype = "dashed", color = "black", aes(group = 1)) + 
    scale_color_gradientn(colors = rainbow(5)) + 
    theme(legend.position = "bottom") + 
    labs(x = "Follow Up Period", 
         y = "Antisocial Score", 
         color = "Home Cog Score"), 

  ggplot(data = curran, 
         mapping = aes(x = log_anti, group = homecog, fill = homecog)) + 
    geom_histogram() + 
    scale_fill_gradientn(colors = rainbow(5)) , 
  
  nrow = 1
)

```


```{r}

#| fig-height: 8
#| fig-width: 12
grid.arrange(
  ggplot(data = curran, 
       aes(x = homecog, y = log_anti)) + 
  geom_point(aes(color = as.factor(occasion))) + 
  geom_smooth() + 
  theme(legend.position = "bottom"), 
  
  ggplot(data = curran, 
       aes(x = homeemo, y = log_anti)) + 
  geom_point(aes(color = as.factor(occasion))) + 
  geom_smooth() + 
  theme(legend.position = "bottom"),
  
  nrow = 1
)
```

\newpage

# Conclusions 

* Might want to use mixed effects model to account for a large amount of individual variation 

* `Homecog` and `Homeemo` might have a weak effect, but after accounting for individual random effect, they might be significant 
after all 

* No visual evidence that we need to consider higher order predictors, no apparent non-linear trends 

* Variance increases as time increases, we will consider a poisson, or some other model to account for this effect of variance 

* Antisocial scores vary quite a lot in the population

* As HOME variables increase, antisocial score decreases 

* Antisocial score fluctuates quite a lot, however, it seems that on average it is higher for kids with lower HOME variables scores
